fGFPCA dynamic prediction (In-sample)

### 2023-01-06 f-GFPCA.Rmd
out-of-sample prediction using Empirical Bayes

@@

Now let's say if we have a new subject $\boldsymbol{Y}_{new}$ with observations up to $t_m$ and $t_m$ is in mth bin. Also, we do not wish to refit any model, including the local GLMMs. In the first step, we need to use the series of local GLMMs that's already been fitted to get the random intercepts for this new observations ($\boldsymbol{b}_{new}$).

We first focus on one specific interval s.

@@

```{r}
```

>>>>

# Rationale

We assume that a binary function $Y(t)$ is generated by a latent Gaussian function $\eta(t)$, as follows:

$$Y_i(t) \sim Binomial(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))})$$
$$\eta_i(t) = \beta_0(t)+b_i(t)$$

And this latent funciton can be approximated by fPCA:

$$\eta_i(t) = f_0(t)+\sum_{k=1}^{K}\xi_k\phi_k(t)+\epsilon_i(t)$$

@@

Now let's say if we have a new subject $\boldsymbol{Y}$ with observations up to $t_m$ and $t_m$ is in mth bin. I can think of a few ways to  predict the following track, as below:

### Refit local GLMMs to estimate latent function

@@

### Do not estimate latent function

Let say on the binned grid, at each interval s, new sample has $n_s$ observations, among which $m_s$ are successes. $m_s = \sum_{j=1}^{n_S} I(Y_{sj}=1)$, j is the index for observation in a specific interval.

We use the expectation of latent function conditonal on fPCA score to estimate the true latent function, so that the error term can be removed from following calculation:

$$\hat{\eta_s} = \hat{f}_{0}(s)+\phi(s)\boldsymbol{\xi}$$

Here, $\hat{f}_{0}(s)$ is the mean function from fPCA and $\phi(s)$ are eigenfunctions at time s, which is a K $\times$ 1 vector.

Then the data is generated from a binomial distribution:

$$\begin{aligned}
 \boldsymbol{Y}_s & \sim Binomial(p_s) \\
 logit(p_s) & = \hat{\eta}_s = \hat{f}_{0}(s)+\phi(s)\boldsymbol{\xi}
\end{aligned}$$

Now let's find the empirical bayes estimator of score $E(\boldsymbol{\xi}|\boldsymbol{Y})$, which is not likely to have closed form solution.

$$E(\boldsymbol{\xi}|\boldsymbol{Y})=\int_{\boldsymbol{\xi}} \boldsymbol{\xi} p(\boldsymbol{\xi}|\boldsymbol{Y})d\boldsymbol{\xi}$$
$$p(\boldsymbol{\xi}|\boldsymbol{Y}) = \frac{p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})}{\int_{\boldsymbol{\xi}}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$

Thus:

$$E(\boldsymbol{\xi}|\boldsymbol{Y})=\frac{\int_{\boldsymbol{\xi}}\boldsymbol{\xi}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}{\int_{\boldsymbol{\xi}}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$

We know that fPCA scores follow a multivariate normal distribution:

$$p(\boldsymbol{\xi}) = (2\pi)^{-K/2}det(\boldsymbol{\Gamma})^{-1/2}exp(-\boldsymbol{\xi^T \Gamma^{-1} \xi}/2)$$

And the binary outcome, conditional on score, follows a Binomial distribution:

$$\begin{aligned}
p(\boldsymbol{Y}|\boldsymbol{\xi}) &= \prod_{s=1}^{m}p(\boldsymbol{Y}_t|\boldsymbol{\xi})
\end{aligned}$$
<!-- We can utilize the fitted series of local GLMMs, instead of refitting them. Since $\boldsymbol{Y}_{new}$ is a binary function, we can get its (incomplete) latent function value $\eta^s$ with Maximum Likelihood Estimation. Specifically, we need to estimate its (incomplete) random intercept function $b^s$. -->

<!-- Let's focus on a specific bin $s$. If the new sample has $n_s$ observations in this bin, among which $m_s$ are success (value = 1). Then $\boldsymbol{Y}^s$ would follow a Binomial distribution:  -->

<!-- $$\boldsymbol{Y}^s \sim Binomial(n_s, \frac{exp(\eta^s)}{1+exp(\eta^s)})$$ -->
<!-- Where $\eta^s = \beta_0^s+b^s$, and $\beta_0^s$ is already estimated with training data.  -->

<!-- Now we write out the likelihood in interval s: -->

<!-- $$L(b^s|\boldsymbol{Y}^s) = \binom{n^s}{m^s}p_s^{m^s}(1-p_s)^{n^s-m^s}$$ -->
<!-- where  -->

<!-- $$p_s = \frac{exp(\eta^s)}{1+exp(\eta^s)}, \hspace{0.5cm}  -->
<!-- \eta^s = \beta_0^s+b^s$$ -->

<!-- Then the log likelihood is: -->

<!-- $$\begin{aligned} -->
<!-- l(b^s|\boldsymbol{Y}^s) &= log\binom{n^s}{m^s}+{m^s}logp_s+(n^s-m^s)log(1-p_s)\\ -->
<!-- &= log\binom{n^s}{m^s}+m^slog(\frac{p_s}{1-p_s})+n^slog(1-p_s)\\ -->
<!-- &= log\binom{n^s}{m^s}+m^s(\beta_0+b^s)-n^slog(1+exp(\beta_0^s+b^s)) -->
<!-- \end{aligned}$$ -->

<!-- Then we take the first derivative and set it to zero, and end up with MLE os $b^s$: -->

<!-- $$\hat{b}^s=log(\frac{m^s}{n^s-m^s})-\beta_0^s$$ -->
<!-- Again, $\beta_0^s$ is the estimated mean latent function from the local GlMMs steps. In fact, this individual latent track can be simply estimated by success ratio in the corresponding interval: -->

<!-- $$\eta^s = log\frac{m^s}{n^s-m^s}$$ -->

<!-- ```{r} -->
<!-- # generate one subject from the smae generating mechanism -->
<!-- xi_new <- matrix(rnorm(1*K),1,K); -->
<!-- xi_new <- xi_new %*% diag(sqrt(lambda)) -->
<!-- # random intercept function -->
<!-- b_new <- xi_new %*% t(phi) -->
<!-- # latent function -->
<!-- eta_new <- f_0(sind) + b_new -->
<!-- ## response binary function -->
<!-- Y_new <- rbinom(length(eta_new), size=1, prob=plogis(eta_new)) -->
<!-- df_new_true <- data.frame( -->
<!--   id = N+1000, -->
<!--   sind = sind,  -->
<!--   Y = Y_new, -->
<!--   eta = t(eta_new), -->
<!--   sind_inx = 1:J) -->

<!-- # cut observations after specific time point -->
<!-- tm <- 0.5 -->
<!-- df_new_obs <- df_new_true %>%  select(-eta) %>%  -->
<!--   filter(sind<=0.5) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # bin observations -->
<!-- df_new_obs$bin <- cut(df_new_obs$sind_inx, breaks = brks,  -->
<!--                       include.lowest = T,labels = mid) -->
<!-- df_new_obs$bin <- as.numeric(as.character(df_new_obs$bin)) -->

<!-- # estimated MLE of latent function -->
<!-- eta_hat_new <- df_new_obs %>% group_by(bin) %>%  -->
<!--   summarise_at("Y", function(x){log(sum(x==1)/sum(x==0))})  -->

<!-- # plot -->
<!-- df_new_true %>% filter(sind_inx %in% mid) %>%  -->
<!--   select(sind_inx, eta) %>%  -->
<!--   rename(bin = sind_inx) %>%  -->
<!--   left_join(eta_hat_new) %>%  -->
<!--   ggplot()+ -->
<!--   geom_line(aes(x=bin, y=eta, col = "True"))+ -->
<!--   geom_line(aes(x=bin, y=Y, col = "Predicted")) -->

<!-- ``` -->
